Place of death in NRS COVID-19 deaths: visualisation for 2022 data

### 2022-06-27 covid_nrs_place_of_death_visualisations_2022.Rmd
tidied up visualisations a bit, and changed direction of colour scale so that high cases/home deaths = dark; also changed the parallel maps to have same colour coding

## Map-based visualisations

@@

### Home deaths increase since 2019 by LA

@@

```{r}
(fig_map_home_deaths_increase_from_2020_la <-
  merged_deaths_la_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority", width=40),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

@@

### Home deaths increase since 2019 by HB

```{r}
(fig_map_home_deaths_increase_from_2020_hb <-
  merged_deaths_hb_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_hb_home_deaths-1, maximum_ratio_between_hb_and_hb_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board", width=40),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

@@

### Home deaths increase since 2019 by LA, by year

@@

```{r}
(fig_map_home_deaths_increase_by_year_la <-
  merged_deaths_la_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(year, ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    # legend.position = "bottom"
    ) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority",
       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

@@

### Home deaths increase vs. covid cases by LA, all deaths since 2019

```{r}
(fig_map_covid_cases_from_2020_la <-
  weekly_covid_cases_la %>%
  group_by(ref_area) %>%
  summarise(
    ratio_covid_cases = max(crude_rate_positive_per100k),
    .groups = "drop"
  ) %>% 
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020, by Local Authority", width=40),
       subtitle = "",
       caption = paste0(c(only_2022_text,crude_rate_text,source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)

@@

### Home deaths increase since 2019 by HB, by year

@@

```{r}
(fig_map_home_deaths_increase_by_year_hb <-
  merged_deaths_hb_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(year, ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_hb_home_deaths-1, maximum_ratio_between_hb_and_hb_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    aspect.ratio = 1.4
    ) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board",
       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

@@

### Home deaths increase vs. covid cases by HB, all deaths since 2019

```{r}
(fig_map_covid_cases_from_2020_hb <-
  weekly_covid_cases_hb %>%
  group_by(ref_area) %>%
  summarise(
    ratio_covid_cases = max(crude_rate_positive_per100k),
    .groups = "drop"
  ) %>% 
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    aspect.ratio = 1.4
    ) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020, by NHS Health Board", width=40),
       subtitle = "",
       caption = paste0(c(only_2022_text,crude_rate_text,source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)

@@

#### LA covid rates over time

@@

```{r}
## the problem with this plot is that there were between 500 per 100k cases to 25k per 100k cases over time and over LAs - how do we visualise that?
(fig_map_covid_cases_by_year_la <-
  weekly_covid_cases_la %>%
  left_join(population_estimate_la_2019 %>% rename(population_estimate = all_ages, year_estimate = year), by = c("ref_area"="la")) %>%
  group_by(ref_area, year, population_estimate) %>%
  summarise(
    cumulative_cases = sum(weekly_positive, na.rm=TRUE),
    ratio_covid_cases = cumulative_cases / population_estimate * 1e5,
    .groups = "drop"
  ) %>%
  distinct %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  # ggplot(aes(geometry=geometry, fill=log(ratio_covid_cases)), colour="grey10") +
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(option = "B", labels = function(x) format(exp(x), big.mark=",")) +
  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020 per 100,000 population, by Local Authority", width=50),
       subtitle = "",
       caption = paste0(c(source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)
```

## cumulative deaths at home over time

>>>>

# Map-based visualisations

@@

## Home deaths increase since 2019 by LA

@@

```{r}
(fig_map_home_deaths_increase_from_2020_la <-
  merged_deaths_la_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority", width=40),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

@@

## Home deaths increase since 2019 by HB

```{r}
(fig_map_home_deaths_increase_from_2020_hb <-
  merged_deaths_hb_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_hb_home_deaths-1, maximum_ratio_between_hb_and_hb_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board", width=40),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

@@

## Home deaths increase since 2019 by LA, by year

@@

```{r}
(fig_map_home_deaths_increase_by_year_la <-
  merged_deaths_la_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(year, ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    ) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority",
       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

@@

## Home deaths increase vs. covid cases by LA, all deaths since 2019

```{r}
(fig_map_covid_cases_from_2020_la <-
  weekly_covid_cases_la %>%
  group_by(ref_area) %>%
  summarise(
    ratio_covid_cases = max(crude_rate_positive_per100k),
    .groups = "drop"
  ) %>% 
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020, by Local Authority", width=40),
       subtitle = "",
       caption = paste0(c(only_2022_text,crude_rate_text,source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)

@@

## Home deaths increase since 2019 by HB, by year

@@

```{r}
(fig_map_home_deaths_increase_by_year_hb <-
  merged_deaths_hb_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(year, ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    aspect.ratio = 1.4
    ) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board",
       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

@@

## Home deaths increase vs. covid cases by HB, all deaths since 2019

```{r}
(fig_map_covid_cases_from_2020_hb <-
  weekly_covid_cases_hb %>%
  group_by(ref_area) %>%
  summarise(
    ratio_covid_cases = max(crude_rate_positive_per100k),
    .groups = "drop"
  ) %>% 
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    aspect.ratio = 1.4
    ) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020, by NHS Health Board", width=40),
       subtitle = "",
       caption = paste0(c(only_2022_text,crude_rate_text,source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)

@@

### LA covid rates over time

@@

```{r}
## the problem with this plot is that there were between 500 per 100k cases to 25k per 100k cases over time and over LAs - how do we visualise that?
(fig_map_covid_cases_by_year_la <-
  weekly_covid_cases_la %>%
  left_join(population_estimate_la_2019 %>% rename(population_estimate = all_ages, year_estimate = year), by = c("ref_area"="la")) %>%
  group_by(ref_area, year, population_estimate) %>%
  summarise(
    cumulative_cases = sum(weekly_positive, na.rm=TRUE),
    ratio_covid_cases = cumulative_cases / population_estimate * 1e5,
    .groups = "drop"
  ) %>%
  distinct %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020 per 100,000 population, by Local Authority", width=50),
       subtitle = "",
       caption = paste0(c(source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)
```

# Cumulative deaths at home over time
